Bayesian Inference in Numerical Cognition: A Tutorial Using JASP

### 2019-12-10 jncTutorial/manuscript.Rmd
add marginal posterior plot

The questions we would like to address here are (1) "Which out of these two covariates (\( x_{1}=\text{ symbolic distance effect} \) and \( x_{2}=\text{ nonsymbolic distance effect}\)) predict fluency score?", and (2) "How large is the effect \( \beta_i \) of each covariate \( x_{i} \)?". Note that the first question can be answered by (a) neither symbolic nor nonsymbolic, (b) only symbolic, (c) only nonsymbolic, and (d) both symbolic and nonsymbolic distance effects predict fluency score. Equivalently, these answers correspond to the hypotheses that the regression coefficients \( \beta_{i} \) (a') for both symbolic and nonsymbolic are equal to 0; (b') for nonsymbolic is equal to zero; (c') for symbolic is equal to zero; and (d') for neither is equal to zero. Hence, the first question is the testing problem, whereas the second question deals with the estimation problem.

@@

Once we have positive evidence for \( \mathcal{M}_{1} \), which posits only symbolicDE as a predictor of fluency, we can move to the problem of estimation. To this end, we focus on the "Posterior Summary" output of Figure \@ref(fig:regression1). This table provides model averaged posterior summaries of the regression coefficients \( \beta_{0} \) (the intercept), \( \beta_{1} \) (the coefficient of symbolicDE), and \( \beta_{2} \) (the coefficient of nonsymbolicDE). Before interpreting the output, however, there are some subtle details that we should discuss. As mentioned earlier in the \( t \)-test example, any estimation that we present is conditional on *some* model. However, in this example, we have four models to consider. So, how to proceed? One approach is to employ Bayesian model averaging, which JASP performs by default. Basically, it works as follows. Rather than estimating a parameters conditioned on one specific model, JASP calculates an *unconditional* posterior distribution for the parameter. That is, we factor in both the uncertainty about the parameter within any one particular model *and* the uncertainty about the models themselves. We think this is a good approach to estimation, as it alleviates the need for the researcher to pick one model as "the winner", and instead, provides estimates that incorporate both levels of uncertainty (within models and across models).

Given this background, let us interpret the Posterior Summary table in Table \@ref(fig:regression1). First, the table provides estimates of mean and standard deviation for the three coefficients, as well as 95% credible intervals. Moreover, the table provides "inclusion probabilities", which provide prior and posterior probabilties of including each predictor as a component of the linear model. We can take a look at these in more detail. First, let us focus on \( P(\text{incl}) \), the prior inclusion probability for each predictor. These are based on the prior model probabilities that we chose in Table \@ref(tabLinreg). Since the intercept \( \beta_{0} \) was included in each of the 4 models, and each model \( \mathcal{M}_{k} \) had prior probability 0.25, the prior probability of including the intercept \( \beta_{0} \) is equal to \( 4(0.25) = 1 \). On the other hand, only two of the models (\( \mathcal{M}_{1} \) and \( \mathcal{M}_{3} \)) included the predictor symbolicDE. Since each of these models had prior probability 0.25, the prior probability of including the predictor symbolicDE is equal to \(2(0.25) = 0.5\). A similar argument shows the same prior probability for the predictor nonsymbolicDE.

@@

>>>>

The questions we would like to address here are (1) "Which out of these two covariates (\( x_{1}=\text{ symbolic distance effect} \) and \( x_{2}=\text{ nonsymbolic distance effect}\)) predict fluency score?", and (2) "How large is the effect \( \beta_i \) of each covariate \( x_{i} \)?". Note that the first question can be answered as: (a) neither symbolic nor nonsymbolic, (b) only symbolic, (c) only nonsymbolic, and (d) both symbolic and nonsymbolic distance effects predict fluency score. Equivalently, these answers correspond to specific hypotheses about the regression coefficients \( \beta_{i} \), namely: (a') the coefficients for both symbolic and nonsymbolic are equal to 0; (b') only the coefficient for nonsymbolic is equal to zero; (c') only the coefficient for symbolic is equal to zero; and (d') neither coefficient is equal to zero. Cast in this manner, the first question is the testing problem, whereas the second question deals with the estimation problem.

@@

Once we have positive evidence for \( \mathcal{M}_{1} \), which posits only symbolicDE as a predictor of fluency, we can move to the problem of estimation. To this end, we focus on two outputs: (1) the "Posterior Summary" output of Figure \@ref(fig:regression1); and (2) the "Marginal posterior distributions" plot (see Figure \@ref(fig:marginalPosterior)), which can be displayed by clicking on the "Plots" menu in JASP. Both outputs provide model averaged posterior summaries of the regression coefficients \( \beta_{0} \) (the intercept), \( \beta_{1} \) (the coefficient of symbolicDE), and \( \beta_{2} \) (the coefficient of nonsymbolicDE). Before interpreting these outputs, however, there are some subtle details that we should discuss. As mentioned earlier in the \( t \)-test example, any estimation that we present is conditional on *some* model. However, in this example, we have four models to consider. So, how to proceed? One approach is to employ Bayesian model averaging, which JASP performs by default. Basically, it works as follows. Rather than estimating a parameters conditioned on one specific model, JASP calculates an *unconditional* posterior distribution for the parameter. That is, we factor in both the uncertainty about the parameter within any one particular model *and* the uncertainty about the models themselves. We think this is a good approach to estimation, as it alleviates the need for the researcher to pick one model as "the winner", and instead, provides estimates that incorporate both levels of uncertainty (within models and across models).

![\label{fig:marginalPosterior}JASP screenshot showing the marginal posterior distribution for the nonsymbolicDE predictor in our linear regression example.](figures/marginalPosterior.png){width=60%}

Given this background, let us first interpret the Posterior Summary table in Table \@ref(fig:regression1). First, the table provides estimates of mean and standard deviation for the three coefficients, as well as 95% credible intervals. Moreover, the table provides "inclusion probabilities", which provide prior and posterior probabilties of including each predictor as a component of the linear model. We can take a look at these in more detail. First, let us focus on \( P(\text{incl}) \), the prior inclusion probability for each predictor. These are based on the prior model probabilities that we chose in Table \@ref(tabLinreg). Since the intercept \( \beta_{0} \) was included in each of the 4 models, and each model \( \mathcal{M}_{k} \) had prior probability 0.25, the prior probability of including the intercept \( \beta_{0} \) is equal to \( 4(0.25) = 1 \). On the other hand, only two of the models (\( \mathcal{M}_{1} \) and \( \mathcal{M}_{3} \)) included the predictor symbolicDE. Since each of these models had prior probability 0.25, the prior probability of including the predictor symbolicDE is equal to \(2(0.25) = 0.5\). A similar argument shows the same prior probability for the predictor nonsymbolicDE.

@@

Finally, let us consider the marginal posterior distribution displayed in Figure \@ref{fig:marginalPosterior}. For simplicity of discussion, we only display the marginal posterior for the nonsymbolicDE predictor, though JASP provides a plot for each predictor being considered. The figure shows exactly what is meant by model averaging; from the discussion above, we know there is uncertainty about whether to include nonsymbolicDE as a predictor in the overall model. This uncertainty can be quantified as the posterior inclusion probability, which from the posterior summary table we know to be \( P(\text{incl}\mid \text{data}) = 0.448 \). Thus, it easily follows that the posterior probability of *excluding* nonsymbolicDE as a predictor is \( 1-0.448 = 0.552 \); this probability is displayed in Figure \@ref(fig:marginalPosterior) as a "spike" at 0. The rest of the distribution displays the probable values of the coefficient for nonsymbolicDE, conditioned on the inclusion of nonsymbolicDE as a predictor. That is, *if* we're willing to accept the idea that nonsymbolicDE is a predictor of fluency, there is a 95% probability that the coefficient for nonsymbolicDE ranges from -6.28 to 35.28.

### 2019-12-10 jncTutorial/manuscript.tex
add marginal posterior plot

The questions we would like to address here are (1) ``Which out of these two covariates (\(x_{1}=\text{ symbolic distance effect}\) and \(x_{2}=\text{ nonsymbolic distance effect}\)) predict fluency score?'', and (2) ``How large is the effect \(\beta_i\) of each covariate \(x_{i}\)?''. Note that the first question can be answered by (a) neither symbolic nor nonsymbolic, (b) only symbolic, (c) only nonsymbolic, and (d) both symbolic and nonsymbolic distance effects predict fluency score. Equivalently, these answers correspond to the hypotheses that the regression coefficients \(\beta_{i}\) (a') for both symbolic and nonsymbolic are equal to 0; (b') for nonsymbolic is equal to zero; (c') for symbolic is equal to zero; and (d') for neither is equal to zero. Hence, the first question is the testing problem, whereas the second question deals with the estimation problem.

@@

Once we have positive evidence for \(\mathcal{M}_{1}\), which posits only symbolicDE as a predictor of fluency, we can move to the problem of estimation. To this end, we focus on the ``Posterior Summary'' output of Figure \ref{fig:regression1}. This table provides model averaged posterior summaries of the regression coefficients \(\beta_{0}\) (the intercept), \(\beta_{1}\) (the coefficient of symbolicDE), and \(\beta_{2}\) (the coefficient of nonsymbolicDE). Before interpreting the output, however, there are some subtle details that we should discuss. As mentioned earlier in the \(t\)-test example, any estimation that we present is conditional on \emph{some} model. However, in this example, we have four models to consider. So, how to proceed? One approach is to employ Bayesian model averaging, which JASP performs by default. Basically, it works as follows. Rather than estimating a parameters conditioned on one specific model, JASP calculates an \emph{unconditional} posterior distribution for the parameter. That is, we factor in both the uncertainty about the parameter within any one particular model \emph{and} the uncertainty about the models themselves. We think this is a good approach to estimation, as it alleviates the need for the researcher to pick one model as ``the winner'', and instead, provides estimates that incorporate both levels of uncertainty (within models and across models).

Given this background, let us interpret the Posterior Summary table in \ref{fig:regression1}. First, the table provides estimates of mean and standard deviation for the three coefficients, as well as 95\% credible intervals. Moreover, the table provides ``inclusion probabilities'', which provide prior and posterior probabilties of including each predictor as a component of the linear model. We can take a look at these in more detail. First, let us focus on \(P(\text{incl})\), the prior inclusion probability for each predictor. These are based on the prior model probabilities that we chose in Table \ref{tabLinreg}. Since the intercept \(\beta_{0}\) was included in each of the 4 models, and each model \(\mathcal{M}_{k}\) had prior probability 0.25, the prior probability of including the intercept \(\beta_{0}\) is equal to \(4(0.25) = 1\). On the other hand, only two of the models (\(\mathcal{M}_{1}\) and \(\mathcal{M}_{3}\)) included the predictor symbolicDE. Since each of these models had prior probability 0.25, the prior probability of including the predictor symbolicDE is equal to \(2(0.25) = 0.5\). A similar argument shows the same prior probability for the predictor nonsymbolicDE.

@@

>>>>

The questions we would like to address here are (1) ``Which out of these two covariates (\(x_{1}=\text{ symbolic distance effect}\) and \(x_{2}=\text{ nonsymbolic distance effect}\)) predict fluency score?'', and (2) ``How large is the effect \(\beta_i\) of each covariate \(x_{i}\)?''. Note that the first question can be answered as: (a) neither symbolic nor nonsymbolic, (b) only symbolic, (c) only nonsymbolic, and (d) both symbolic and nonsymbolic distance effects predict fluency score. Equivalently, these answers correspond to specific hypotheses about the regression coefficients \(\beta_{i}\), namely: (a') the coefficients for both symbolic and nonsymbolic are equal to 0; (b') only the coefficient for nonsymbolic is equal to zero; (c') only the coefficient for symbolic is equal to zero; and (d') neither coefficient is equal to zero. Cast in this manner, the first question is the testing problem, whereas the second question deals with the estimation problem.

@@

Once we have positive evidence for \(\mathcal{M}_{1}\), which posits only symbolicDE as a predictor of fluency, we can move to the problem of estimation. To this end, we focus on two outputs: (1) the ``Posterior Summary'' output of Figure \ref{fig:regression1}; and (2) the ``Marginal posterior distributions'' plot (see Figure \ref{fig:marginalPosterior}), which can be displayed by clicking on the ``Plots'' menu in JASP. Both outputs provide model averaged posterior summaries of the regression coefficients \(\beta_{0}\) (the intercept), \(\beta_{1}\) (the coefficient of symbolicDE), and \(\beta_{2}\) (the coefficient of nonsymbolicDE). Before interpreting these outputs, however, there are some subtle details that we should discuss. As mentioned earlier in the \(t\)-test example, any estimation that we present is conditional on \emph{some} model. However, in this example, we have four models to consider. So, how to proceed? One approach is to employ Bayesian model averaging, which JASP performs by default. Basically, it works as follows. Rather than estimating a parameters conditioned on one specific model, JASP calculates an \emph{unconditional} posterior distribution for the parameter. That is, we factor in both the uncertainty about the parameter within any one particular model \emph{and} the uncertainty about the models themselves. We think this is a good approach to estimation, as it alleviates the need for the researcher to pick one model as ``the winner'', and instead, provides estimates that incorporate both levels of uncertainty (within models and across models).

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth,height=\textheight]{figures/marginalPosterior.png}
\caption{\label{fig:marginalPosterior}JASP screenshot showing the marginal posterior distribution for the nonsymbolicDE predictor in our linear regression example.}
\end{figure}

Given this background, let us first interpret the Posterior Summary table in Table \ref{fig:regression1}. First, the table provides estimates of mean and standard deviation for the three coefficients, as well as 95\% credible intervals. Moreover, the table provides ``inclusion probabilities'', which provide prior and posterior probabilties of including each predictor as a component of the linear model. We can take a look at these in more detail. First, let us focus on \(P(\text{incl})\), the prior inclusion probability for each predictor. These are based on the prior model probabilities that we chose in Table \ref{tabLinreg}. Since the intercept \(\beta_{0}\) was included in each of the 4 models, and each model \(\mathcal{M}_{k}\) had prior probability 0.25, the prior probability of including the intercept \(\beta_{0}\) is equal to \(4(0.25) = 1\). On the other hand, only two of the models (\(\mathcal{M}_{1}\) and \(\mathcal{M}_{3}\)) included the predictor symbolicDE. Since each of these models had prior probability 0.25, the prior probability of including the predictor symbolicDE is equal to \(2(0.25) = 0.5\). A similar argument shows the same prior probability for the predictor nonsymbolicDE.

@@

Finally, let us consider the marginal posterior distribution displayed in Figure @ref\{fig:marginalPosterior\}. For simplicity of discussion, we only display the marginal posterior for the nonsymbolicDE predictor, though JASP provides a plot for each predictor being considered. The figure shows exactly what is meant by model averaging; from the discussion above, we know there is uncertainty about whether to include nonsymbolicDE as a predictor in the overall model. This uncertainty can be quantified as the posterior inclusion probability, which from the posterior summary table we know to be \(P(\text{incl}\mid \text{data}) = 0.448\). Thus, it easily follows that the posterior probability of \emph{excluding} nonsymbolicDE as a predictor is \(1-0.448 = 0.552\); this probability is displayed in Figure \ref{fig:marginalPosterior} as a ``spike'' at 0. The rest of the distribution displays the probable values of the coefficient for nonsymbolicDE, conditioned on the inclusion of nonsymbolicDE as a predictor. That is, \emph{if} we're willing to accept the idea that nonsymbolicDE is a predictor of fluency, there is a 95\% probability that the coefficient for nonsymbolicDE ranges from -6.28 to 35.28.
